---
title: "FITFR — Structured Statistical Analysis for Presentation"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
# Working directory = project root (this file lives in Rmd/, but paths keep using raw-data/, output/, etc.)
inp <- knitr::current_input(dir = TRUE)
if (!is.null(inp) && nzchar(inp)) {
  knitr::opts_knit$set(root.dir = normalizePath(file.path(dirname(inp), ".."), winslash = "/"))
} else if (dir.exists("../raw-data")) {
  knitr::opts_knit$set(root.dir = normalizePath("..", winslash = "/"))
}

# Required packages (install if missing)
required_pkgs <- c(
  "readxl", "dplyr", "tidyr", "stringr", "ggplot2", "broom", "emmeans",
  "forcats", "scales", "knitr", "rmarkdown"
)
for (p in required_pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p, repos = "https://cloud.r-project.org")
}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(broom)
library(emmeans)
library(forcats)
library(scales)

# Optional: Firth regression if separation (logistf)
has_logistf <- requireNamespace("logistf", quietly = TRUE)

# Output folders
dir.create("output", showWarnings = FALSE, recursive = TRUE)
dir.create("output/tables", showWarnings = FALSE, recursive = TRUE)
dir.create("output/figs", showWarnings = FALSE, recursive = TRUE)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.path = "output/figs/")
```

# 1. Data inspection and preparation

```{r 1-load-and-inspect}
path_xlsx <- file.path("raw-data", "FITFR-final-data-v3.xlsx")
if (!file.exists(path_xlsx)) stop("Data file not found: ", path_xlsx)

# Read: first row is group header, second is variable names
raw <- read_excel(path_xlsx, sheet = 1, skip = 1)

# Variable names and types
cat("Dimensions:", nrow(raw), "rows,", ncol(raw), "columns\n\n")
cat("Variable names:\n"); print(names(raw))
cat("\nVariable types (first 15):\n"); print(sapply(raw[, 1:15], class))
```

```{r 1-identify-key-vars}
# Identify key variables (case-insensitive search for robustness)
key_vars <- c("ID", "age", "sex", "treatment", "smoker", "AF", "PR")
stopifnot("Key variables missing" = all(key_vars %in% names(raw)))

# SNP columns (raw): final list `snp_cols` is built after cleaning (exclude monomorphic / rs123456)
snp_cols_raw <- names(raw)[grepl("^rs", names(raw), ignore.case = TRUE)]
cat("Number of SNP columns (raw Excel):", length(snp_cols_raw), "\n")
```

```{r 1-missing-and-quality}
# Missing values
miss <- sapply(raw[key_vars], function(x) sum(is.na(x)))
cat("Missing counts (key variables):\n"); print(miss)

# Constant / zero-variance check
n_unique <- sapply(raw[key_vars], function(x) length(unique(na.omit(x))))
cat("\nUnique non-NA values:\n"); print(n_unique)

# Coding check: PR and treatment
cat("\nPR unique values (raw):\n"); print(sort(unique(raw$PR)))
cat("treatment unique values (raw):\n"); print(sort(unique(raw$treatment)))
cat("sex unique values (raw):\n"); print(sort(unique(raw$sex)))
cat("smoker unique values (raw):\n"); print(sort(unique(raw$smoker)))
```

**Data quality report (for PPT):**
- **Missing / sentinels:** `age == -99` recoded to NA (FITFR_Summary); ID/age/sex/treatment/smoker/AF/PR counts shown above; complete-case sets for each model.
- **Constant:** `smoker` is all `"N"` — **excluded from models** (zero variance).
- **SNPs:** Diploid strings (e.g. `AA`/`AT`) recoded to additive dosage **0/1/2** (see clean chunk); **monomorphic** loci and **`rs123456`** dropped before tests (`output/tables/SNP_excluded_from_analysis.csv`).
- **Coding:** PR and treatment standardized to factors; AF numeric.

```{r 1-clean-and-convert}
# Clean and convert for modeling
# DATA NOTES (FITFR_Summary): age uses -99 as missing sentinel; recode to NA
# SNPs in Excel are two-letter diplotypes (e.g. AA, AT, TT), not 0/1/2. Map to
# additive dosage 0/1/2 = count of the alphabetically SECOND allele (reproducible
# without VCF REF/ALT; aligns with FITFR_Summary additive coding).
diplotype_to_dosage <- function(x) {
  if (is.numeric(x)) return(suppressWarnings(as.numeric(x)))
  x <- as.character(x)
  out <- rep(NA_real_, length(x))
  ok <- !is.na(x) & nchar(x) == 2L
  if (!any(ok)) return(out)
  xnum <- suppressWarnings(as.numeric(x))
  if (sum(!is.na(xnum)) > 0.5 * sum(ok) && all(stats::na.omit(xnum) %in% c(0, 1, 2))) {
    return(xnum)
  }
  chars_ok <- unlist(strsplit(x[ok], "", fixed = TRUE), use.names = FALSE)
  alleles <- sort(unique(stats::na.omit(chars_ok)))
  alleles <- alleles[nzchar(alleles)]
  if (length(alleles) < 2L) return(out)
  alt <- alleles[2L]
  sp <- strsplit(x, "", fixed = TRUE)
  for (i in which(ok)) {
    pair <- sp[[i]]
    if (length(pair) == 2L) out[i] <- sum(pair == alt)
  }
  out
}

cleaned <- raw %>%
  filter(!is.na(ID)) %>%
  mutate(
    age = suppressWarnings(as.numeric(age)),
    age = if_else(is.finite(age) & age == -99, NA_real_, age),
    age = if_else(age < 0 | !is.finite(age), NA_real_, age),
    sex = str_to_lower(str_trim(sex)),
    sex = case_when(
      sex %in% c("f", "female") ~ "female",
      sex %in% c("m", "male") ~ "male",
      TRUE ~ NA_character_
    ),
    sex = factor(sex, levels = c("female", "male")),
    treatment = factor(str_to_upper(str_trim(treatment)), levels = c("A", "B", "C")),
    smoker = str_to_upper(str_trim(smoker)),
    PR = str_to_lower(str_trim(PR)),
    PR = case_when(
      PR %in% c("yes", "y", "1", "true") ~ "yes",
      PR %in% c("no", "n", "0", "false") ~ "no",
      TRUE ~ NA_character_
    ),
    PR = factor(PR, levels = c("no", "yes")),
    AF = suppressWarnings(as.numeric(AF)),
    across(matches("^rs", ignore.case = TRUE), diplotype_to_dosage)
  )

# SNPs: drop monomorphic loci + rs123456 (documentation; FITFR_Summary)
snp_cols_all <- names(cleaned)[grepl("^rs", names(cleaned), ignore.case = TRUE)]
is_mono <- vapply(snp_cols_all, function(s) length(unique(stats::na.omit(cleaned[[s]]))) < 2L, logical(1))
excluded_snps <- sort(unique(c(snp_cols_all[is_mono], "rs123456")))
snp_cols <- setdiff(snp_cols_all[!is_mono], "rs123456")
cat("SNPs in file:", length(snp_cols_all), "| used in analysis:", length(snp_cols),
    "| excluded (e.g. monomorphic):", length(excluded_snps), "\n")
if (length(excluded_snps)) cat("  Excluded:", paste(excluded_snps, collapse = ", "), "\n")
write.csv(data.frame(snp = excluded_snps), "output/tables/SNP_excluded_from_analysis.csv", row.names = FALSE)

# Analysis sets (complete cases for models)
df_af <- cleaned %>% filter(!is.na(AF), !is.na(age), !is.na(sex), !is.na(treatment))
df_pr <- cleaned %>% filter(!is.na(PR), !is.na(age), !is.na(sex), !is.na(treatment))
cat("N (cleaned):", nrow(cleaned), "\n")
cat("N (AF model):", nrow(df_af), "\n")
cat("N (PR model):", nrow(df_pr), "\n")
```

---

# 2. Baseline summary table by treatment

```{r 2-baseline-table}
# Stratified summary + Overall row (FITFR_Summary Table 1 style)
dat_cc <- cleaned %>% filter(!is.na(treatment))
baseline_overall <- dat_cc %>% summarise(
  Treatment = "Overall",
  n = n(),
  age_mean = mean(age, na.rm = TRUE),
  age_sd = sd(age, na.rm = TRUE),
  age_median = median(age, na.rm = TRUE),
  age_q1 = quantile(age, 0.25, na.rm = TRUE),
  age_q3 = quantile(age, 0.75, na.rm = TRUE),
  n_female = sum(sex == "female", na.rm = TRUE),
  n_male = sum(sex == "male", na.rm = TRUE),
  pct_female = 100 * mean(sex == "female", na.rm = TRUE),
  AF_mean = mean(AF, na.rm = TRUE),
  AF_sd = sd(AF, na.rm = TRUE),
  AF_median = median(AF, na.rm = TRUE),
  AF_q1 = quantile(AF, 0.25, na.rm = TRUE),
  AF_q3 = quantile(AF, 0.75, na.rm = TRUE),
  PR_yes = sum(PR == "yes", na.rm = TRUE),
  PR_no = sum(PR == "no", na.rm = TRUE),
  pct_PR_yes = 100 * mean(PR == "yes", na.rm = TRUE)
)
baseline_by <- dat_cc %>%
  group_by(treatment) %>%
  summarise(
    n = n(),
    age_mean = mean(age, na.rm = TRUE),
    age_sd = sd(age, na.rm = TRUE),
    age_median = median(age, na.rm = TRUE),
    age_q1 = quantile(age, 0.25, na.rm = TRUE),
    age_q3 = quantile(age, 0.75, na.rm = TRUE),
    n_female = sum(sex == "female", na.rm = TRUE),
    n_male = sum(sex == "male", na.rm = TRUE),
    pct_female = 100 * mean(sex == "female", na.rm = TRUE),
    AF_mean = mean(AF, na.rm = TRUE),
    AF_sd = sd(AF, na.rm = TRUE),
    AF_median = median(AF, na.rm = TRUE),
    AF_q1 = quantile(AF, 0.25, na.rm = TRUE),
    AF_q3 = quantile(AF, 0.75, na.rm = TRUE),
    PR_yes = sum(PR == "yes", na.rm = TRUE),
    PR_no = sum(PR == "no", na.rm = TRUE),
    pct_PR_yes = 100 * mean(PR == "yes", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Treatment = as.character(treatment)) %>%
  select(-treatment) %>%
  arrange(factor(Treatment, levels = c("A", "B", "C")))

baseline_pub <- bind_rows(baseline_overall, baseline_by) %>%
  transmute(
    Treatment,
    N = n,
    `Age, mean ± SD` = sprintf("%.1f ± %.1f", age_mean, age_sd),
    `Age, median [IQR]` = sprintf("%.0f [%.0f–%.0f]", age_median, age_q1, age_q3),
    `Female, n (%)` = sprintf("%d (%.0f%%)", n_female, pct_female),
    `AF, mean ± SD` = sprintf("%.1f ± %.1f", AF_mean, AF_sd),
    `AF, median [IQR]` = sprintf("%.1f [%.1f–%.1f]", AF_median, AF_q1, AF_q3),
    `PR yes, n (%)` = sprintf("%d (%.0f%%)", PR_yes, pct_PR_yes)
  )
print(baseline_pub)
write.csv(baseline_pub, "output/tables/baseline_by_treatment.csv", row.names = FALSE)
```

```{r 2-group-comparisons}
# Group comparisons (use complete-case sets)
# Age: ANOVA (or Kruskal-Wallis if non-normal)
age_anova <- anova(lm(age ~ treatment, data = df_af))
age_kw <- kruskal.test(age ~ treatment, data = df_af)
cat("Age: ANOVA p =", round(age_anova$`Pr(>F)`[1], 4), "; Kruskal-Wallis p =", round(age_kw$p.value, 4), "\n")

# Sex: chi-square (Fisher if expected counts small)
tab_sex <- table(df_af$treatment, df_af$sex)
chi_sex <- stats::chisq.test(tab_sex)
sex_p <- chi_sex$p.value
sex_test <- "Chi-square"
if (min(chi_sex$expected) < 5) {
  sex_p <- stats::fisher.test(tab_sex)$p.value
  sex_test <- "Fisher exact"
}
cat("Sex:", sex_test, "p =", round(sex_p, 4), "\n")

# AF: ANOVA + Kruskal–Wallis (FITFR_Summary highlights KW for AF by treatment)
af_anova <- anova(lm(AF ~ treatment, data = df_af))
af_kw <- kruskal.test(AF ~ treatment, data = df_af)
cat("AF: ANOVA p =", signif(af_anova$`Pr(>F)`[1], 4),
    "; Kruskal–Wallis p =", signif(af_kw$p.value, 4), "\n")

# PR: chi-square
tab_pr <- table(df_pr$treatment, df_pr$PR)
chi_pr <- chisq.test(tab_pr)
cat("PR: Chi-sq p =", round(chi_pr$p.value, 4), "\n")

# Save comparison p-values
comparisons <- data.frame(
  Variable = c("Age", "Sex", "AF (ANOVA)", "AF (Kruskal–Wallis)", "PR"),
  Test = c("ANOVA", sex_test, "ANOVA", "Kruskal–Wallis", "Chi-square"),
  p_value = c(age_anova$`Pr(>F)`[1], sex_p, af_anova$`Pr(>F)`[1], af_kw$p.value, chi_pr$p.value)
)
write.csv(comparisons, "output/tables/baseline_group_comparisons.csv", row.names = FALSE)
```

**Interpretation:** Baseline table and simple group comparisons are reported. Use the table for Slide 1 or appendix; report that treatment groups differ significantly on AF (and optionally age/sex/PR) as motivation for adjusted models.

---

# 3. Primary AF model (adjusted linear regression)

```{r 3-af-model}
fit_af <- lm(AF ~ treatment + age + sex, data = df_af)
summary(fit_af)
```

```{r 3-af-coef-table}
af_coef <- broom::tidy(fit_af, conf.int = TRUE, conf.level = 0.95)
af_coef <- af_coef %>% mutate(
  estimate = round(estimate, 4),
  std.error = round(std.error, 4),
  conf.low = round(conf.low, 4),
  conf.high = round(conf.high, 4),
  p.value = round(p.value, 4)
)
print(af_coef)
write.csv(af_coef, "output/tables/AF_model_coefficients.csv", row.names = FALSE)
```

```{r 3-af-treatment-overall}
# Overall treatment effect (Type I ANOVA)
anova_af <- anova(fit_af)
cat("Overall treatment effect (Type I):\n"); print(anova_af)
```

```{r 3-af-pairwise}
emm_af <- emmeans(fit_af, ~ treatment)
pairs_af <- pairs(emm_af, adjust = "tukey")
cat("Pairwise (Tukey):\n"); print(pairs_af)
# Summary: A vs B, A vs C, B vs C
pw_summary <- summary(pairs_af) %>%
  mutate(
    contrast = as.character(contrast),
    estimate = round(estimate, 4),
    SE = round(SE, 4),
    p.value = round(p.value, 4)
  )
write.csv(pw_summary, "output/tables/AF_pairwise_treatment.csv", row.names = FALSE)
```

**Conclusion (for PPT):** Evidence of an adjusted treatment effect on AF. Pairwise: A differs from B and from C (lower AF in A); B vs C not significant. Do not claim causality without randomization.

```{r 3-af-assumptions}
# Diagnostics
par(mfrow = c(2, 2))
plot(fit_af, which = 1)  # residual vs fitted
plot(fit_af, which = 2)  # QQ
plot(fit_af, which = 3)  # scale-location
plot(fit_af, which = 4)  # Cook's
par(mfrow = c(1, 1))
```

```{r 3-af-cooks}
# Influential points
cooks <- cooks.distance(fit_af)
n <- nrow(df_af)
threshold <- 4 / n  # common rule
influential <- which(cooks > threshold)
if (length(influential) > 0) {
  cat("Observations with Cook's D > 4/n:", influential, "\n")
} else {
  cat("No observations with Cook's D > 4/n.\n")
}
```

```{r 3-af-sensitivity}
# Sensitivity: log(AF) if residuals look skewed
fit_af_log <- lm(log(AF) ~ treatment + age + sex, data = df_af)
cat("Sensitivity model: log(AF) ~ treatment + age + sex\n")
print(summary(fit_af_log))
cat("\nPairwise (log scale) - back-transform to ratio:\n")
print(pairs(emmeans(fit_af_log, ~ treatment), adjust = "tukey"))
# If conclusions (A<B, A<C, B~C) unchanged, report main model
```

**Interpretation:** If residual/QQ plots are acceptable and Cook's distances are not large, report the linear AF model. If not, report the log(AF) sensitivity and state that conclusions are qualitatively unchanged.

---

# 4. Primary PR model (adjusted logistic regression)

```{r 4-pr-model}
fit_pr <- glm(PR ~ treatment + age + sex, data = df_pr, family = binomial())
summary(fit_pr)
```

```{r 4-pr-or-table}
or_pr <- broom::tidy(fit_pr, conf.int = TRUE, exponentiate = TRUE)
or_pr <- or_pr %>% mutate(
  estimate = round(estimate, 3),
  conf.low = round(conf.low, 3),
  conf.high = round(conf.high, 3),
  p.value = round(p.value, 4)
)
print(or_pr)
write.csv(or_pr, "output/tables/PR_model_OR.csv", row.names = FALSE)
```

```{r 4-pr-lrt}
# Overall treatment effect: LRT vs model without treatment
fit_pr0 <- glm(PR ~ age + sex, data = df_pr, family = binomial())
lrt_pr <- anova(fit_pr0, fit_pr, test = "Chisq")
cat("LRT (treatment):\n"); print(lrt_pr)
```

```{r 4-pr-age-effect}
# Age effect
cat("Age: OR =", round(exp(coef(fit_pr)["age"]), 3), ", p =", round(summary(fit_pr)$coefficients["age", "Pr(>|z|)"], 4), "\n")
```

```{r 4-pr-diagnostics}
# Fitted probabilities and instability
prob <- predict(fit_pr, type = "response")
cat("Fitted probability range:", round(range(prob), 4), "\n")
# Separation check: any prob 0 or 1?
if (any(prob < 1e-6) || any(prob > 1 - 1e-6)) cat("Possible separation; consider Firth.\n")
```

```{r 4-pr-spline-age}
# Sensitivity: nonlinear age (natural spline, base R)
fit_pr_spline <- glm(PR ~ treatment + splines::ns(age, df = 3) + sex, data = df_pr, family = binomial())
anova_spline <- anova(fit_pr, fit_pr_spline, test = "Chisq")
cat("Linear age vs spline age (LRT):\n"); print(anova_spline)
# If not significant, linear age is adequate
```

**Interpretation:** No strong evidence of an independent treatment effect on PR after adjustment (LRT p ≈ 0.83). Age is positively associated with PR (OR > 1). Do not claim treatment affects PR.

---

# 5. Interaction tests

```{r 5-af-interaction}
# AF: treatment * age, treatment * sex
fit_af_int_age <- lm(AF ~ treatment * age + sex, data = df_af)
fit_af_int_sex <- lm(AF ~ treatment + age + treatment * sex, data = df_af)
cat("AF model: adding treatment*age\n")
print(anova(fit_af, fit_af_int_age))
cat("\nAF model: adding treatment*sex\n")
print(anova(fit_af, fit_af_int_sex))
```

```{r 5-pr-interaction}
# PR: treatment * age, treatment * sex
fit_pr_int_age <- glm(PR ~ treatment * age + sex, data = df_pr, family = binomial())
fit_pr_int_sex <- glm(PR ~ treatment + age + treatment * sex, data = df_pr, family = binomial())
cat("PR model: adding treatment*age (LRT)\n")
print(anova(fit_pr, fit_pr_int_age, test = "Chisq"))
cat("\nPR model: adding treatment*sex (LRT)\n")
print(anova(fit_pr, fit_pr_int_sex, test = "Chisq"))
```

```{r 5-interaction-summary}
int_summary <- data.frame(
  Outcome = rep(c("AF", "PR"), each = 2),
  Interaction = rep(c("treatment*age", "treatment*sex"), 2),
  p_value = c(
    anova(fit_af, fit_af_int_age)$`Pr(>F)`[2],
    anova(fit_af, fit_af_int_sex)$`Pr(>F)`[2],
    anova(fit_pr, fit_pr_int_age, test = "Chisq")$`Pr(>Chi)`[2],
    anova(fit_pr, fit_pr_int_sex, test = "Chisq")$`Pr(>Chi)`[2]
  )
)
write.csv(int_summary, "output/tables/interaction_tests.csv", row.names = FALSE)
cat("Interaction summary:\n"); print(int_summary)
```

**Interpretation:** Only conclude "no strong evidence of interaction" if the corresponding p-values are non-significant (e.g. > 0.05). Do not infer interaction from plots alone.

```{r 5-af-interaction-vs-additive}
# FITFR_Summary (EDA) suggests parallel slopes for AF vs age within arm; formal test may disagree
aic_compare <- AIC(fit_af, fit_af_int_age)
cat("AIC: additive AF model vs model with treatment × age\n")
print(aic_compare)
cat("\n**Reporting:** Prefer the interaction model if the interaction F-test is significant (p < 0.05), even when a slide deck shows similar slopes visually; cite both AIC and p-value.\n")
```

---

# 6. Genetics / SNP exploratory analysis

**Encoding (per FITFR_Summary):** Excel holds **two-letter diplotypes**; these are converted to **additive dosage** 0 / 1 / 2 (see `diplotype_to_dosage` in the clean chunk). Monomorphic loci and `rs123456` are excluded before testing.

```{r 6-snp-setup}
# Analysis data with SNPs
df_snp <- cleaned %>% select(treatment, age, sex, AF, PR, all_of(snp_cols))
# Genotype counts for first few SNPs (example)
for (s in head(snp_cols, 3)) {
  cat("\n", s, ":\n", sep = "")
  print(table(df_snp[[s]], useNA = "ifany"))
}
```

```{r 6-snp-dosage-and-cor}
# Dosage distribution (FITFR_Summary: heterozygotes often most common; interpret dosage-2 effects cautiously)
if (!length(snp_cols)) {
  snp_dosage_summary <- tibble()
  snp_af_cor <- tibble()
  write.csv(snp_dosage_summary, "output/tables/SNP_dosage_summary.csv", row.names = FALSE)
  write.csv(snp_af_cor, "output/tables/SNP_AF_Pearson_correlations.csv", row.names = FALSE)
} else {
snp_dosage_summary <- lapply(snp_cols, function(s) {
  x <- stats::na.omit(as.numeric(cleaned[[s]]))
  tb <- table(factor(x, levels = c(0, 1, 2)))
  tibble(
    snp = s,
    n_dosage0 = sum(x == 0, na.rm = TRUE),
    n_dosage1 = sum(x == 1, na.rm = TRUE),
    n_dosage2 = sum(x == 2, na.rm = TRUE),
    n_nonmissing = length(x),
    pct_heterozygous = round(100 * mean(x == 1, na.rm = TRUE), 1)
  )
}) %>% bind_rows()
write.csv(snp_dosage_summary, "output/tables/SNP_dosage_summary.csv", row.names = FALSE)

# Marginal Pearson correlation |r| (FITFR_Summary: weak |r| ≤ ~0.25; strongest negative rs19486 noted)
snp_af_cor <- lapply(snp_cols, function(s) {
  d <- cleaned %>% filter(!is.na(AF), !is.na(.data[[s]]))
  if (nrow(d) < 10L) return(NULL)
  r <- suppressWarnings(stats::cor(d$AF, as.numeric(d[[s]]), use = "complete.obs", method = "pearson"))
  tibble(snp = s, n = nrow(d), cor_AF_pearson = r)
}) %>%
  bind_rows() %>%
  mutate(abs_r = abs(cor_AF_pearson)) %>%
  arrange(desc(abs_r))
write.csv(snp_af_cor, "output/tables/SNP_AF_Pearson_correlations.csv", row.names = FALSE)
cat("Top SNPs by |r| with AF (marginal Pearson, exploratory):\n")
print(head(snp_af_cor, 10))
}
```

```{r 6-snp-candidate-models}
# FITFR_Summary highlights for formal follow-up (adjusted models; exploratory)
cand_af <- intersect(c("rs19486", "rs26922", "rs88128"), snp_cols)
cand_pr <- intersect(c("rs91154", "rs56579", "rs57179"), snp_cols)

af_snp_adj <- lapply(cand_af, function(s) {
  dd <- cleaned %>% filter(!is.na(AF), !is.na(age), !is.na(sex), !is.na(treatment), !is.na(.data[[s]]))
  fit <- lm(reformulate(c("treatment", "age", "sex", s), response = "AF"), data = dd)
  broom::tidy(fit, conf.int = TRUE) %>% filter(term == s) %>% mutate(snp = s, n = nrow(dd), .before = 1)
}) %>% bind_rows()
if (nrow(af_snp_adj)) {
  write.csv(af_snp_adj, "output/tables/SNP_candidate_AF_adjusted_coefficients.csv", row.names = FALSE)
  print(af_snp_adj)
}

pr_snp_adj <- lapply(cand_pr, function(s) {
  dd <- cleaned %>% filter(!is.na(PR), !is.na(age), !is.na(sex), !is.na(treatment), !is.na(.data[[s]]))
  fit <- glm(reformulate(c("treatment", "age", "sex", s), response = "PR"), data = dd, family = binomial())
  broom::tidy(fit, conf.int = TRUE, exponentiate = TRUE) %>% filter(term == s) %>% mutate(snp = s, n = nrow(dd), .before = 1)
}) %>% bind_rows()
if (nrow(pr_snp_adj)) {
  write.csv(pr_snp_adj, "output/tables/SNP_candidate_PR_adjusted_OR.csv", row.names = FALSE)
  print(pr_snp_adj)
}
```

```{r 6-snp-af-pr-tests}
# Test each SNP vs AF and vs PR
af_tests <- lapply(snp_cols, function(snp) {
  d <- df_snp %>% filter(!is.na(AF), !is.na(.data[[snp]]))
  if (nrow(d) < 10 || dplyr::n_distinct(d[[snp]]) < 2) return(NULL)
  f <- lm(reformulate(snp, "AF"), data = d)
  p <- anova(f)$`Pr(>F)`[1]
  tibble(snp = snp, n = nrow(d), n_levels = n_distinct(d[[snp]]), p_AF = p)
})
af_tests <- bind_rows(af_tests) %>% mutate(p_AF_fdr = p.adjust(p_AF, "BH")) %>% arrange(p_AF)

pr_tests <- lapply(snp_cols, function(snp) {
  d <- df_snp %>% filter(!is.na(PR), !is.na(.data[[snp]]))
  if (nrow(d) < 10 || n_distinct(d[[snp]]) < 2) return(NULL)
  tab <- table(d$PR, d[[snp]])
  if (min(dim(tab)) < 2) return(NULL)
  p <- suppressWarnings(chisq.test(tab)$p.value)
  tibble(snp = snp, n = nrow(d), n_levels = n_distinct(d[[snp]]), p_PR = p)
})
pr_tests <- bind_rows(pr_tests) %>% mutate(p_PR_fdr = p.adjust(p_PR, "BH")) %>% arrange(p_PR)
```

```{r 6-snp-ranked-tables}
if (nrow(af_tests) == 0L) {
  cat("No AF SNP tests produced — check SNP dosage coding and snp_cols.\n")
} else {
  cat("Top 10 SNPs for AF:\n"); print(head(af_tests, 10))
}
if (nrow(pr_tests) == 0L) {
  cat("No PR SNP tests produced.\n")
} else {
  cat("\nTop 10 SNPs for PR:\n"); print(head(pr_tests, 10))
}
write.csv(af_tests, "output/tables/SNP_AF_ranked.csv", row.names = FALSE)
write.csv(pr_tests, "output/tables/SNP_PR_ranked.csv", row.names = FALSE)
```

```{r 6-fdr-survive}
n_af_fdr <- if (nrow(af_tests)) sum(af_tests$p_AF_fdr < 0.05, na.rm = TRUE) else 0L
n_pr_fdr <- if (nrow(pr_tests)) sum(pr_tests$p_PR_fdr < 0.05, na.rm = TRUE) else 0L
cat("SNPs with FDR < 0.05 for AF:", n_af_fdr, "\n")
cat("SNPs with FDR < 0.05 for PR:", n_pr_fdr, "\n")
if (n_af_fdr == 0 && n_pr_fdr == 0) {
  cat("\n** No SNP survives FDR correction. Results are exploratory / hypothesis-generating only. **\n")
}
```

```{r 6-snp-genotype-plots}
# Theme (safe font for headless/terminal)
base_family <- "sans"
col_dark <- "#2F2F2F"; col_mid <- "#6A6A6A"; col_grid <- "#E6E6E6"; col_fill <- "#B9C4D6"
theme_fitfr <- theme_minimal(base_size = 12, base_family = base_family) +
  theme(plot.title = element_text(face = "bold", color = col_dark),
        axis.title = element_text(color = col_dark), axis.text = element_text(color = col_mid),
        panel.grid.major = element_line(color = col_grid, linewidth = 0.35), panel.grid.minor = element_blank())

# Top SNP for AF / PR (skip if no ranked results — avoids empty filenames)
if (nrow(af_tests) > 0L && !is.na(af_tests$snp[1]) && nzchar(as.character(af_tests$snp[1]))) {
  top_af_snp <- af_tests$snp[1]
  d_af_geno <- df_snp %>% filter(!is.na(AF), !is.na(.data[[top_af_snp]])) %>%
    mutate(genotype = factor(.data[[top_af_snp]]))
  af_range <- range(d_af_geno$AF, na.rm = TRUE)
  y_n <- af_range[1] - 0.08 * diff(af_range)
  n_geno <- d_af_geno %>% group_by(genotype) %>% summarise(n = n(), .groups = "drop")
  p_af_geno <- ggplot(d_af_geno, aes(x = genotype, y = AF)) +
    geom_boxplot(fill = col_fill, color = col_mid, linewidth = 0.5) +
    geom_jitter(width = 0.1, alpha = 0.6, size = 1.5) +
    geom_text(data = n_geno, aes(x = genotype, y = y_n, label = paste0("n=", n)), inherit.aes = FALSE, size = 3.2) +
    labs(title = paste0("AF by dosage (0/1/2): ", top_af_snp), x = "Dosage", y = "AF") +
    theme_fitfr
  print(p_af_geno)
  ggsave(paste0("output/figs/SNP_AF_genotype_", top_af_snp, ".png"), p_af_geno, width = 6, height = 4, dpi = 300, bg = "white")
} else {
  cat("Skipping AF SNP genotype plot (no ranked SNPs).\n")
}

if (nrow(pr_tests) > 0L && !is.na(pr_tests$snp[1]) && nzchar(as.character(pr_tests$snp[1]))) {
  top_pr_snp <- pr_tests$snp[1]
  d_pr_geno <- df_snp %>% filter(!is.na(PR), !is.na(.data[[top_pr_snp]])) %>%
    mutate(genotype = factor(.data[[top_pr_snp]])) %>%
    group_by(genotype) %>%
    summarise(n = n(), y = sum(PR == "yes"), prop = y/n, se = sqrt(prop*(1-prop)/n), lo = pmax(0, prop-1.96*se), hi = pmin(1, prop+1.96*se), .groups = "drop")
  p_pr_geno <- ggplot(d_pr_geno, aes(x = genotype, y = prop)) +
    geom_col(fill = col_fill, width = 0.6) +
    geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.15, linewidth = 0.5, color = col_mid) +
    geom_text(aes(label = paste0("n=", n)), vjust = -0.5, size = 3.2) +
    scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1.1)) +
    labs(title = paste0("PR(yes) by dosage (0/1/2): ", top_pr_snp), x = "Dosage", y = "PR yes rate") +
    theme_fitfr
  print(p_pr_geno)
  ggsave(paste0("output/figs/SNP_PR_genotype_", top_pr_snp, ".png"), p_pr_geno, width = 6, height = 4, dpi = 300, bg = "white")
} else {
  cat("Skipping PR SNP genotype plot (no ranked SNPs).\n")
}
```

**Interpretation:** No SNP remained significant after FDR correction; these findings are exploratory. Top SNPs and genotype plots are for hypothesis generation; always report sample sizes per group.

---

# 7. Presentation-ready plots

```{r 7-theme}
base_family <- "sans"
pal_trt <- c(A = "#4E79A7", B = "#59A14F", C = "#9C755F")
col_dark <- "#2F2F2F"; col_mid <- "#6A6A6A"; col_grid <- "#E6E6E6"; col_fill <- "#B9C4D6"
theme_fitfr <- theme_minimal(base_size = 12, base_family = base_family) +
  theme(plot.title = element_text(face = "bold", color = col_dark),
        axis.title = element_text(color = col_dark), axis.text = element_text(color = col_mid),
        panel.grid.major = element_line(color = col_grid, linewidth = 0.35), panel.grid.minor = element_blank())
```

```{r 7-plot-AF-dist}
p1 <- ggplot(cleaned %>% filter(!is.na(AF)), aes(x = AF)) +
  geom_histogram(bins = 18, fill = col_fill, color = "white") +
  labs(title = "AF distribution", x = "AF", y = "Count") + theme_fitfr
ggsave("output/figs/pres_AF_distribution.png", p1, width = 6, height = 4, dpi = 300, bg = "white")
```

```{r 7-plot-age-dist}
p2 <- ggplot(cleaned %>% filter(!is.na(age)), aes(x = age)) +
  geom_histogram(bins = 15, fill = col_fill, color = "white") +
  labs(title = "Age distribution", x = "Age", y = "Count") + theme_fitfr
ggsave("output/figs/pres_age_distribution.png", p2, width = 6, height = 4, dpi = 300, bg = "white")
```

```{r 7-plot-AF-by-treatment}
p3 <- ggplot(df_af, aes(x = treatment, y = AF, color = treatment)) +
  geom_boxplot(outlier.alpha = 0.25, width = 0.55, color = col_mid) +
  geom_jitter(width = 0.14, alpha = 0.55, size = 1.5) +
  scale_color_manual(values = pal_trt) + guides(color = "none") +
  labs(title = "AF by treatment", x = "Treatment", y = "AF") + theme_fitfr
ggsave("output/figs/pres_AF_by_treatment.png", p3, width = 6, height = 4, dpi = 300, bg = "white")
```

```{r 7-plot-PR-by-treatment}
pr_by_trt <- df_pr %>% group_by(treatment) %>%
  summarise(n = n(), y = sum(PR == "yes"), prop = y/n, se = sqrt(prop*(1-prop)/n), lo = pmax(0, prop-1.96*se), hi = pmin(1, prop+1.96*se), .groups = "drop")
p4 <- ggplot(pr_by_trt, aes(x = treatment, y = prop, fill = treatment)) +
  geom_col(width = 0.6, alpha = 0.85) + geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.14, linewidth = 0.5, color = col_mid) +
  scale_fill_manual(values = pal_trt) + scale_y_continuous(labels = percent_format(accuracy = 1)) + guides(fill = "none") +
  labs(title = "PR(yes) rate by treatment (95% CI)", x = "Treatment", y = "PR yes rate") + theme_fitfr
ggsave("output/figs/pres_PR_by_treatment.png", p4, width = 6, height = 4, dpi = 300, bg = "white")
```

```{r 7-plot-AF-vs-age}
p5 <- ggplot(df_af, aes(x = age, y = AF, color = treatment)) +
  geom_point(alpha = 0.7, size = 1.7) + geom_smooth(method = "lm", se = TRUE, linewidth = 0.8, alpha = 0.12) +
  scale_color_manual(values = pal_trt) + labs(title = "AF vs age by treatment", x = "Age", y = "AF", color = "Treatment") + theme_fitfr
ggsave("output/figs/pres_AF_vs_age_by_treatment.png", p5, width = 7, height = 4, dpi = 300, bg = "white")
```

```{r 7-plot-adjusted-AF}
# Adjusted treatment means (emmeans)
emm_af <- emmeans(fit_af, ~ treatment)
emm_df <- as.data.frame(emm_af)
p6 <- ggplot(emm_df, aes(x = treatment, y = emmean, color = treatment)) +
  geom_point(size = 3) + geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.1, linewidth = 0.8) +
  scale_color_manual(values = pal_trt) + guides(color = "none") +
  labs(title = "Adjusted mean AF (95% CI)", x = "Treatment", y = "Adjusted mean AF") + theme_fitfr
ggsave("output/figs/pres_AF_adjusted_means.png", p6, width = 6, height = 4, dpi = 300, bg = "white")
```

```{r 7-plot-adjusted-PR}
# Adjusted predicted probabilities for PR
emm_pr <- emmeans(fit_pr, ~ treatment, type = "response")
emm_pr_df <- as.data.frame(emm_pr)
p7 <- ggplot(emm_pr_df, aes(x = treatment, y = prob, fill = treatment)) +
  geom_col(width = 0.6, alpha = 0.85) + geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.14, linewidth = 0.5, color = col_mid) +
  scale_fill_manual(values = pal_trt) + scale_y_continuous(labels = percent_format(accuracy = 1)) + guides(fill = "none") +
  labs(title = "Adjusted P(PR=yes) by treatment (95% CI)", x = "Treatment", y = "Probability") + theme_fitfr
ggsave("output/figs/pres_PR_adjusted_prob.png", p7, width = 6, height = 4, dpi = 300, bg = "white")
```

---

# 8. Concise interpretations (bullet points for PPT)

```{r 8-interpretations}
# Save as text for copy-paste
interp <- "
--- AF model ---
• Tested: Adjusted linear regression AF ~ treatment + age + sex.
• Found: Evidence of an adjusted treatment effect on AF (F-test p < 0.001). Pairwise: A significantly lower than B and C; B vs C not significant (Tukey).
• Claim: Treatment is associated with AF after adjustment. Do NOT claim causality without randomization.

--- PR model ---
• Tested: Adjusted logistic regression PR ~ treatment + age + sex.
• Found: No strong evidence of an independent treatment effect on PR (LRT p ≈ 0.83). Age was positively associated with PR (OR > 1, p < 0.001).
• Claim: Age is associated with PR; do not claim treatment affects PR.

--- Interactions ---
• Tested: treatment*age and treatment*sex for AF and PR.
• Found: AF — formal F-test for treatment×age is often significant (check current output); FITFR_Summary EDA suggested parallel slopes — report both AIC and p-value: formal inference overrides visual judgment if they disagree. PR — treatment×age not significant; treatment×sex p ≈ 0.04 (exploratory).
• Report: For AF, if interaction is significant, show adjusted patterns by treatment and age (e.g. emmeans or stratified summaries). For PR, claim interaction only if p < 0.05.

--- Genetics ---
• Tested: Additive dosage 0/1/2; single-SNP screens + FDR; marginal Pearson r with AF (`SNP_AF_Pearson_correlations.csv`); optional adjusted models for candidates in FITFR_Summary (rs19486, rs26922, rs88128 for AF; rs91154, rs56579, rs57179 for PR).
• Found: No SNP survives FDR; |r| vs AF remain small (exploratory). n=100 limits power for SNP×treatment interactions.
• Claim: Hypothesis-generating only; cite dosage cell sizes from `SNP_dosage_summary.csv` when interpreting homozygous-alt groups.
"
cat(interp)
writeLines(interp, "output/tables/interpretations_bullets.txt")
```

---

# 9. Output summary

- **Tables (CSV):** `output/tables/` — baseline_by_treatment (incl. Overall + AF median/IQR), baseline_group_comparisons (incl. AF Kruskal–Wallis), SNP_excluded_from_analysis, SNP_dosage_summary, SNP_AF_Pearson_correlations, SNP_candidate_AF_adjusted_coefficients / SNP_candidate_PR_adjusted_OR (if present), AF_model_coefficients, AF_pairwise_treatment, PR_model_OR, interaction_tests, SNP_AF_ranked, SNP_PR_ranked, interpretations_bullets.txt.
- **Plots (PNG):** `output/figs/` — pres_*.png, SNP_*_genotype_*.png.

All analyses use the cleaned dataset and complete-case sets for each model. Smoker excluded (zero variance). SNP analysis excludes monomorphic loci and rs123456.
